---
title: 'Assignment #4'
author: "Roland"
date: "2026-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Data: Retailers' Inventories to Sales Ratio (RETAILIRNSA), U.S. Census Bureau via FRED. Monthly, not seasonally adjusted, January 1992 to July 2026. Each model forecasts the next 12 months.

# Part 1: Code

## 1. Load the data, plot, and ACF

```{r load}
library(fpp2)
library(TTR)

RETAILIRNSA <- read.csv("RETAILIRNSA.csv")
ts_data <- ts(RETAILIRNSA$RETAILIRNSA, start = c(1992, 1), frequency = 12)

attributes(ts_data)
plot(ts_data)
Acf(ts_data)
```

## 2. Mean, Naive, Random Walk, and Seasonal Naive

```{r simple-models}
# Take the mean of all available history
mean_forecast <- meanf(ts_data, 12)
plot(mean_forecast)

# Naive
naive_forecast <- naive(ts_data, 12)
plot(naive_forecast)

# Random walk with drift
rwf_forecast <- rwf(ts_data, 12, drift = TRUE)

# Seasonal naive
snaive_forecast <- snaive(ts_data, 12)
```

## 3. Moving Averages

```{r moving-avg}
MA5_smooth <- ma(ts_data, order = 5)
MA9_smooth <- ma(ts_data, order = 9)

plot(ts_data)
lines(MA5_smooth, col = "red")
lines(MA9_smooth, col = "blue")

# Moving average forecast = average of the last k months, held flat
MA5_forecast <- ts(rep(tail(SMA(ts_data, 5), 1), 12), start = c(2026, 8), frequency = 12)
MA9_forecast <- ts(rep(tail(SMA(ts_data, 9), 1), 12), start = c(2026, 8), frequency = 12)
MA5_forecast[1]
MA9_forecast[1]
```

## 4. ETS

```{r ets}
ets_model <- ets(ts_data)
ets_model
plot(ets_model)
ets_model$mse
```

## 5. Holt-Winters and Simple Smoothing

```{r holtwinters}
HW_model <- HoltWinters(ts_data)
HW_model$alpha
HW_model$beta
HW_model$gamma
HW_model$SSE
plot(HW_model)

SSE_Simple <- HoltWinters(ts_data, beta = FALSE, gamma = FALSE)
SSE_Simple$alpha
SSE_Simple$SSE
head(SSE_Simple$fitted)
plot(SSE_Simple)
```

## 6. Forecast the next 12 months

```{r forecasts}
ets_forecast <- forecast(ets_model, h = 12)
HW_forecast  <- forecast(HW_model, h = 12)
SES_forecast <- forecast(SSE_Simple, h = 12)

plot(ets_forecast)
```

## 7. All forecasts in one chart

```{r all-forecasts, fig.width=11, fig.height=6.5}
plot(window(ts_data, start = c(2023, 1)), xlim = c(2023, 2027.7),
     ylim = c(0.95, 1.75), lwd = 3,
     main = "Actual and Model Forecasts (Aug 2026 - Jul 2027)",
     ylab = "Inventory / Sales Ratio")
lines(mean_forecast$mean,   col = "gray50",  lwd = 2, lty = 3)
lines(naive_forecast$mean,  col = "red",     lwd = 2)
lines(rwf_forecast$mean,    col = "green3",  lwd = 2, lty = 2)
lines(snaive_forecast$mean, col = "orange",  lwd = 2)
lines(MA5_forecast,         col = "magenta", lwd = 2, lty = 2)
lines(MA9_forecast,         col = "blue",    lwd = 2, lty = 2)
lines(SES_forecast$mean,    col = "purple",  lwd = 2, lty = 3)
lines(HW_forecast$mean,     col = "brown",   lwd = 2)
lines(ets_forecast$mean,    col = "cyan3",   lwd = 3)
legend("top", ncol = 5, cex = 0.8, bg = "white",
       legend = c("Actual", "Mean", "Naive", "Drift", "Seasonal naive", "MA(5)",
                  "MA(9)", "Simple smoothing", "Holt-Winters", "ETS"),
       col = c("black", "gray50", "red", "green3", "orange", "magenta",
               "blue", "purple", "brown", "cyan3"),
       lwd = c(3, 2, 2, 2, 2, 2, 2, 2, 2, 3), lty = c(1, 3, 1, 2, 1, 2, 2, 3, 1, 1))
```

## 8. Accuracy comparison (RMSE)

```{r accuracy}
results <- c(
  "Mean"             = accuracy(mean_forecast)[1, "RMSE"],
  "Naive"            = accuracy(naive_forecast)[1, "RMSE"],
  "Drift"            = accuracy(rwf_forecast)[1, "RMSE"],
  "Seasonal naive"   = accuracy(snaive_forecast)[1, "RMSE"],
  "MA(5)"            = sqrt(mean(((ts_data - c(NA, head(SMA(ts_data, 5), -1)))[-1])^2, na.rm = TRUE)),
  "MA(9)"            = sqrt(mean(((ts_data - c(NA, head(SMA(ts_data, 9), -1)))[-1])^2, na.rm = TRUE)),
  "Simple smoothing" = accuracy(SES_forecast)[1, "RMSE"],
  "Holt-Winters"     = accuracy(HW_forecast)[1, "RMSE"],
  "ETS"              = accuracy(ets_forecast)[1, "RMSE"]
)
knitr::kable(data.frame(RMSE = round(sort(results), 3)),
             caption = "RMSE of each model, best (lowest) first")
```


